---
title: "Forecasting daily bike rental demand using time series models"
date: "`r Sys.Date()`"
output: html_document
author: "Wellspring Praise"
editor_options: 
  markdown: 
    wrap: 72
---

## About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for
the project on forecasting daily bike rental demand using time series
models in R. It contains analysis such as data exploration, summary
statistics and building the time series models. The final report was
completed on `r date()`.

**Data Description:**

This dataset contains the daily count of rental bike transactions
between years 2011 and 2012 in Capital bikeshare system with the
corresponding weather and seasonal information.

Bike-sharing rental process is highly correlated to the environmental
and seasonal settings. For instance, weather conditions, precipitation,
day of week, season, hour of the day, etc. can affect the rental
behaviors.

Bike sharing systems are new generation of traditional bike rentals
where whole process from membership, rental and return back has become
automatic. Through these systems, user is able to easily rent a bike
from a particular position and return back at another position.
Currently, there are about over 500 bike-sharing programs around the
world which is composed of over 500 thousands bicycles. Today, there
exists great interest in these systems due to their important role in
traffic, environmental and health issues.

More information on the dataset can be gotten from this:

**Data Source:**
<https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset>

**Relevant Paper:**

Fanaee-T, Hadi, and Gama, Joao. Event labeling combining ensemble
detectors and background knowledge, Progress in Artificial Intelligence
(2013): pp. 1-15, Springer Berlin Heidelberg

### Loading data and installing packages

```{r}
## Import required packages
install.packages('stats')
install.packages('forecast')
install.packages('highcharter')
library(stats)
library(forecast)
library(highcharter)
## Loading the datasets
data <- read.csv("hour.csv")
```

```{r}
print(head(data))
```

### Data cleaning

```{r}
## checking quality of data before analysis

# Checking if there are any missing values in the data frame (hour_df and day_df)
cat("Are there any missing data in the dataframe?: ", anyNA(data), "\n")
```

### Describing and exploring the data

```{r}
print(summary(data))
```

This Analysis is centered around the variable 'cnt' which tells us the
total count of registered users per hour(hour_df) or per day(day_df).

**The summary statistic for the 'cnt' variable shows that;**

-   The minimum registered users on a particular day is 1

-   Has an average number of daily registered users of 190.

### Exploring seasonal effects on the Number of Registered Users

-   **How does Holiday affects the number of Registered users?**

```{r}
install.packages("dplyr")
library(dplyr)

holiday_day <- data %>% group_by(holiday) %>% summarize(registered_users = sum(cnt))
holiday_day <- holiday_day %>% mutate(holiday = recode(holiday, `0` = "No", `1` = "Yes"))


print(holiday_day)
```

Lets put this in a visualization to better present it

```{r}
barplot(holiday_day$registered_users, names.arg = holiday_day$holiday, col = "blue", main = "Effects of holiday on number of registered users", xlab = "Holiday", ylab = "Number of registered Users")
```

The seasonality (Holidays) really affects the number of users that
registers. The difference between the Number of Registered Users when
there are no holidays and Number of Registered users when there are
holidays is very large.

The seasonality of holidays, i.e there are holidays, contributes 2.4% to
the number of users registered in that program.

This explains that, during holidays, most percentage of people don't go
out for businesses or work.

-   **How does Seasons(spring, summer, fall, winter) affects the number
    of Registered users?**

Now, lets go over the relationship between the seasons of a year and the
number of registered users. Does the season have any effect on the
number of users registering? well, let's quench the curiosity by
actually looking at how each of them contributes to the number of
registered users.

Since the seasons variable contains discrete values, we would be using
the barplot to show the relationship.

```{r}
## grouping the data 
seasons <- data %>% group_by(season) %>% summarize(registered_users = sum(cnt))
seasons <- seasons %>% mutate(season = recode(season, `1` = "spring", `2` = "summer", `3` = "fall", `4` = "winter"))

max_value = max(seasons$registered_users)

barplot(seasons$registered_users, names.arg = seasons$season, col = "purple", main = "Effects of Seasons on number of  registered users", ylim = c(0, max_value+200000), xlab = "seasons", ylab = "Number of registered Users")

```

During the spring seasons, there are low registrations, and during the
fall season, the number of registration is at its peak than any other
seasons.

Research tells us that spring season is the period after winter, i.e, it
is the transmission period from cold to warm, whereby, fall seasons(also
known as autumn) is the transmission period from warm to cold.

This supports our analysis of number of registered users by seasons. At
the spring season, the temperature is changing from cold to warm, this
indicates that a whole population of people don't ride bikes on a sunny
or hot period, because of health challenges and some other factors.
While, looking at the fall season(autumn), it has the highest number of
registered users, this is because there is a change in the temperature
from warm(hot) to cold.

To put it all together: - Users registers more during a cold season than
hot(warm) seasons. - And, seasons affect the number of registered users.

-   **Relationship between temperature and the number of registered
    users?**

```{r}
plot(x = data$cnt, y = data$temp, main='Relationship Between Temp and Users registration', xlab='No of registered users', ylab='temperature in celscius')

```

The temperature does not neccesarily affect the registration rate of
users. The visualization shows that there is no correlation between the
number of registered users and temperature of that day.

### Creating interactive time series plots

```{r}
# installing packages
install.packages('timetk')
install.packages('plotly')
install.packages('lubridate')

#loading packages
library(plotly)
library(timetk)
library(lubridate)

# getting the time series data
data_time = data[c('dteday', 'cnt')]

# Converting the date variable to a date format
data_time$dteday <- as.Date(data_time$dteday)

# Converting the Date column to a time series object using timetk
data_time <- tk_tbl(data_time, date_var = "dteday")


# Create an interactive time series plot using plotly
plot_ly(data_time, x = ~dteday, y = ~cnt, type = "scatter", mode = "lines") %>%
  layout(title = "Interactive Time Series Plot", xaxis = list(title = "date"), yaxis = list(title = "number of registered users"))
```

### Smoothing time series data

Smoothing time series data involves reducing noise or fluctuations in
the data to reveal underlying trends and patterns more clearly.

There are several methods to smooth time series data. Here we would be
performing the Holt-Winters exponential smoothing.

```{r}
install.packages("forecast")
library(forecast)

# Perform Holt-Winters smoothing
smoothed_data <- ets(data_time$cnt, model = 'ZZZ')

data_time_smooth <- data_time

data_time_smooth$cnt <- fitted(smoothed_data)

plot(data_time$dteday, data_time$cnt, type = 'l', col = 'red', xlab = 'Date', ylab = 'number of registered users')
lines(data_time$dteday, fitted(smoothed_data), type = 'l', col = 'blue')
legend('topleft', legend = c('Original', 'Smoothed'), col = c('red', 'blue'), lty = 1)

```

### Fitting and forecasting time series data using Autoregressive Integrated Moving Average (ARIMA) models

```{r}
install.packages("tseries")
library(tseries)

# Checking for stationarity using the Augmented Dickey-Fuller test
adf_test <- adf.test(data_time_smooth$cnt)

print(adf_test)
```

We see that the p-value is lower than the significance level of 0.05,
thereby rejecting the null hypothesis(not stationary) and accepting the
alternative hypothesis(stationary).

Therefore, we conclude that the data is stationary.

```{r}
# fitting an arima model
arima_model <- auto.arima(data_time_smooth$cnt)

print(arima_model)
```

The auto.arima() function automatically selects the best ARIMA model
order based on AIC (Akaike Information Criterion).

The ARIMA model it selected, ARIMA(2,1,1) indicates:

AR(2): Two autoregressive terms. I(1): The time series has been
differenced one times. MA(1): One moving average terms.

```{r}
#forecasting with the arima model
forecast_values <- forecast(arima_model, h = 31)  # Forecasting 1 month ahead

print(forecast_values, "\n")

```

```{r}
cat("Forcasted number of users that will register in the next 1 month: ", sum(forecast_values$mean))
```

```{r}
plot(forecast_values, main = "Next 1 Month Forecast", ylim = c(0, 2000), xlim = c(16000, dim(data_time_smooth)[1] + 110), xlab = "Day", ylab = "No of registered users")
```

The forecasts are shown as a blue line, with the 80% prediction
intervals as a dark ash shaded area, and the 95% prediction intervals as
an ash shaded area.

### Findings and Conclusions

-   Seasonality affects the rental demand of bikes

-   The Impacts of each seasons of the year on the rental demand of
    bikes are as follows:

    -   Fall(Autumn) seasons contributes **32.2%** to the rental demand
        of bikes.

    -   Summer seasons contributes **27.9%** to the rental demand of
        bikes.

    -   Winter seasons contributes **25.6%** to the rental demand of
        bikes.

    -   Spring Season contributes the lowest, **14.3%**, to the rental
        demand of bikes

-   Holidays has a negative effect on the rental demand of bikes, there
    is a decrease of demand during holidays.

-   The Fall(Autumn) season contributes the highest to the rental
    demands of bikes.

-   There is no effect of temperature to the demand of bikes.

-   The time series model forecasts that there would be a **0.06%**
    increase for the rental demand of bikes within the next 1 month.
